### 2017-01-17 text/Chapter04.tex
Fix error
Edits pages 31-94

In this chapter, we develop a \emph{flat coeffect calculus} that provides a type system for
tracking per-context properties of context-aware programming languages. The \emph{coeffect
calculus} captures the shared properties of such languages. It is parameterized by a \emph{flat
coeffect algebra} and can be instantiated to track implicit parameters, liveness and number of
required past values in dataflow languages.
To capture contextual properties in full generality, the flat coeffect calculus permits multiple
valid typing derivations for a given term. To resolve the ambiguity arising from such generality,
each concrete context-aware language is also equipped with an algorithm for choosing a unique
typing derivation. This allows us to explore the language design landscape, while still follow
the usual scoping rules for languages with established approaches (e.g.~implicit parameters in
Haskell).

@@

\item The coeffect calculus permits multiple typing derivations due to the ambiguity inherent
  in contextual lambda abstraction. Each concrete context-aware language based on the
  coeffect calculus must specify how such ambiguities are to be resolved. We give the procedure for
  choosing unique typing derivation for our three examples (Section~\ref{sec:flat-unique}).

@@

Despite the differences, the systems fit the same framework. This becomes apparent when
we consider the categorical structure (Section~\ref{sec:flat-semantics}). Rather than starting
from the categorical semantics, we first explain how the systems can be unified syntactically
(Section~\ref{sec:flat-calculus-lambda}) and then provide the semantics as an additional
justification.

The development in this chapter can be seen as a counterpart to the well-known development of
\emph{effect systems} \cite{effects-gifford}. Chapter~\ref{ch:semantics} then links \emph{coeffects}
with \emph{comonads} in the same way in which effect systems have been linked with monads
\cite{monad-notions}. The syntax and type system of the flat coeffect calculus follows
a similar style as effect systems \cite{effects-polymorphic,effects-talpin-et-al}, but differs
in the structure of lambda abstraction as discussed briefly here and in Section~\ref{sec:applications-structure-lam}
(the relationship with monads is further discussed in Section~\ref{sec:semantics-related}).

@@

We enrich types and typing judgements with coeffect annotations $\cclrd{r}, \cclrd{s}, \cclrd{t}$;
typing judgements are written as $\coctx{\Gamma}{\cclrd{r}}\vdash e:\tau$. The expressions of
the calculus are those of the $\lambda$-calculus with \emph{let} binding. We also include a type
\ident{num} as an example of a concrete base type with numerical constants written as $n$:
%
\begin{equation*}
\begin{array}{rcl}
e &::=& x \sep n \sep \lambda x:\tau.e \sep e_1~e_2 \sep \kvd{let}~x = e_1~\kvd{in}~e_2\\
\tau &::=& \ident{num} \sep \tau_1 \xrightarrow{\cclrd{r}} \tau_2
\end{array}
\end{equation*}
%
Note that the lambda abstraction in the syntax is written in the Church-style and requires a type
annotation. This will be used in Section~\ref{sec:flat-unique} where we discuss how to find a
unique typing derivation for context-aware computations. Using Church-style lambda abstraction,
we can directly focus on the more interesting problem of finding unique \emph{coeffect annotations}
rather than solving the problem of type reconstruction.

@@

\begin{definition}
A \emph{\cclrd{flat coeffect algebra}} $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$ is a set
$\C$ together with elements $\cunit, \czero \in \C$, binary relation $\cleq$ and binary operations
$\cseq, \cpar, \czip$ such that $(\C, \cseq, \cunit)$ is a monoid, $(\C, \cpar, \czero)$ is an
idempotent monoid, $(\C, \czip)$ is a band (idempotent semigroup) and $(\C, \cleq)$ is a pre-order.
That is, for all $r,s,t\in \C$:
%
\begin{equation*}
\begin{array}{ccc}
r \;\cseq\; (s \;\cseq\; t) = (r \;\cseq\; s) \;\cseq\; t &&
\cunit \;\cseq\; r = r = r \;\cseq\; \cunit
\\
r \;\cpar\; (s \;\cpar\; t) = (r \;\cpar\; s) \;\cpar\; t &
r\; \cpar\; r = r~~~~&
\czero \;\cpar\; r = r = r \;\cpar\; \czero
\\
r \;\czip\; (s \;\czip\; t) = (r \;\czip\; s) \;\czip\; t &&
r\; \czip\; r = r \qquad
\\
\textnormal{if}~~r\; \cleq\; s ~~\textnormal{and}~~s\; \cleq\; t~~\textnormal{then}~~r\; \cleq\; t &&
t\; \cleq\; t \\[0.5em]
\end{array}
\end{equation*}
%
In addition, the following distributivity axioms hold:
\begin{align*}
\quad (\cclrd{r}\, \cpar\, \cclrd{s}) \;\cseq\; \cclrd{t} & = (\cclrd{r} \,\cseq\, \cclrd{t}) \;\cpar\; (\cclrd{s}\, \cseq\, \cclrd{t}) \\
\quad \cclrd{t} \;\cseq\; (\cclrd{r}\, \cpar\, \cclrd{s}) & = (\cclrd{t} \,\cseq\, \cclrd{r}) \;\cpar\; (\cclrd{t}\, \cseq\, \cclrd{s})
\end{align*}
\end{definition}

@@

\begin{itemize}
\item The monoid $(\C, \cseq, \cunit)$ represents \emph{sequential} composition of (semantic)
functions. The monoid axioms are required in order to form a category structure in the
semantics (Section~\ref{sec:flat-semantics}).

@@

\noindent
It is worth noting that, in some of the systems, the operators $\cpar$ and $\czip$ are the least
upper bound and the greatest lower bounds of a lattice. For example, in dataflow computations, they
are \emph{max} and \emph{min} respectively. However, this duality does not hold for implicit parameters
(we discuss the lattice-based formulation of coeffects in Section~\ref{sec:unified-impl-semilattice}.

\paragraph{Ordering.}

The flat coeffect algebra includes a pre-order relation $\cleq$. This will be used to introduce
subcoeffecting and subtyping in Section~\ref{sec:flat-exts-sub}, but we make it a part of the flat
coeffect algebra, as it will be useful for characterization of different kinds of coeffect calculi.
When the idempotent monoid $(\C, \cpar, \czero)$ is also commutative (\ie~forms a
semi-lattice), the $\cleq$ relation can be defined as the ordering of the semi-lattice:
%
\begin{equation*}
\cclrd{r} \;\cleq\; \cclrd{s} \;\Longleftrightarrow\; \cclrd{r} \;\cpar\; \cclrd{s} \;=\; \cclrd{s}
\end{equation*}
%
This definition is consistent with all three examples that motivate flat coeffect calculus, but
it cannot be used with the structural coeffects (Chapter~\ref{ch:structural}), where it fails for
the bounded reuse calculus. For this reason, we choose not to use it for flat coeffect calculus
either.

@@

To guide understanding of the system, we also show non-syntax-directed (\emph{sub}) rule for
subcoeffecting. The rule states that an expression with context demands $\cclrd{r'}$ can be
treated as an expression with greater context demands $\cclrd{r}$. We return to subcoeffecting,
subtyping and additional constructs such as pairs in Section~\ref{sec:flat-exts}. When
discussing procedure for choosing unique typing in Section~\ref{sec:flat-unique}, we consider
only the syntax-directed part of the system.

%---------------------------------------------------------------------------------------------------
s
\subsection{Understanding flat coeffects}
\label{sec:flat-calculus-understanding}

Before proceeding, let us clarify how the typing judgements should be understood. The coeffect
calculus can be understood in two ways discussed in this and the next chapter. As a type system
(Chapter~\ref{ch:flat}), it provides analysis of context dependence. As a semantics
(Chapter~\ref{ch:semantics}), it specifies how context is propagated. These two readings provide
different ways of interpreting the judgements $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ and the
typing rules used to define it.

@@

\noindent
For example, using the syntactic reading, the operators $\czip$ and $\cpar$ represent \emph{merging}
and \emph{splitting} of context demands -- in the (\emph{abs}) rule, $\czip$ appears in the assumption
and the combined context demands of the body are split between two positions in the conclusions;
in the (\emph{app}) rule, $\cpar$ appears in the conclusion and combines two context demands
from the assumptions.

@@

\noindent
For simplicity, assume that all parameters have the same type $\ident{num}$ and so the annotations only
track sets of names. The definition uses a set union for all three operations. Both variables and
constants are are annotated with $\emptyset$ and the ordering is defined by $\subseteq$. The
definition satisfies the flat coeffect algebra axioms because $(S, \cup, \emptyset)$ is an idempotent,
commutative monoid. The language has additional syntax for defining an implicit parameter and for
accessing it, together with associated typing rules:
%
\begin{equation*}
\quad~ e ~::=~ \ldots \sep \ident{?p} \sep \kvd{let}~\ident{?p}~=~e_1~\kvd{in}~e_2
\end{equation*}
\begin{equation*}
\tyrule{param}
  { ~ }
  { \coctx{\Gamma}{\cclrd{ \{ \ident{?p} \} }} \vdash \ident{?p} : \ident{num} }
\end{equation*}
\begin{equation*}
\tyrule{letpar}
  { \coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 &
    \coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_2}
  {\coctx{\Gamma}{\cclrd{r} \;\cclrd{\cup\; (\cclrd{s} \setminus \{ \ident{?p}) \}} } \vdash \kvd{let}~\ident{?p}=e_1~\kvd{in}~e_2 : \tau_2 }
\end{equation*}
%
The (\emph{param}) rule specifies that the accessed parameter $\ident{?p}$ needs to be in the set of
required parameters $\cclrd{r}$. As discussed earlier, we use the same type $\ident{num}$ for all
parameters, but it is also possible to define a coeffect calculus that uses mappings from names to
types (care is needed to avoid assigning multiple types to a parameter of the same type).

@@

As in Section~\ref{sec:applications-flat-live}, sequential composition $\cseq$ is modelled by
the meet operation $\sqcap$ and pointwise composition $\cpar$ is modelled by join $\sqcup$.
The two-point lattice is a commutative, idempotent monoid. Distributivity
$(r \sqcup s) \sqcap t = (r \sqcap t) \sqcup (s \sqcap t)$ does not hold for \emph{every}
lattice, but it trivially holds for the two-point lattice used here.

@@

As required by the axioms, $(\mathbb{N}, +, 0)$ and $(\mathbb{N}, \mathit{max}, 0)$ form monoids
and $(\mathbb{N}, \mathit{min})$ forms a band. Note that dataflow is our first example where
$\cseq$ is not idempotent. The distributivity axioms require the following to be the case:
$\mathit{max}(r,s) + t = \mathit{max}(r+t, s+t)$, which is easy to see.

A simple dataflow language includes an additional construct $\kvd{prev}$ for accessing the
previous value in a stream with an additional typing rule that look as follows:
%
\begin{equation*}
\quad~ e ~::=~ \ldots \sep \kvd{next}~e
\end{equation*}
\begin{equation*}
\tyrule{prev}
  { \coctx{\Gamma}{\cclrd{n}} \vdash e : \tau }
  { \coctx{\Gamma}{\cclrd{n+1}} \vdash \kvd{prev}~e : \tau }
\end{equation*}
%
As a further example that was not covered earlier, it is also possible to combine liveness analysis
and dataflow. In the above dataflow calculus, $0$ denotes that we require the current value of some
variable, but no previous values. However, for constants, we do not even need the current value.

\begin{example}[Optimized dataflow]
In optimized dataflow, context is annotated with natural numbers extended with the $\bot$ element,
that is $\mathbb{N}_{\bot} = \{\bot, 0, 1, 2, 3, \ldots \}$ such that $\forall n \in \mathbb{N}. \bot \leq n$.
The flat coeffect algebra is $(\mathbb{N}_{\bot}, +, \mathit{max}, \mathit{min}, 0, \bot, \leq)$
where $m + n$ is $\bot$ whenever $m=\bot$ or $n=\bot$ and \emph{min}, \emph{max} treat $\bot$ as the
least element.
\end{example}

\noindent
Note that $(\mathbb{N}_{\bot}, +, 0)$ is a monoid for the extended definition of $+$; for the
bottom element $0 + \bot = \bot$ and for natural numbers $0 + n = n$. The structure
$(\mathbb{N}, \emph{max}, \bot)$ is also a monoid, because $\bot$ is the least element and so
$\mathit{max}(n, \bot) = n$. Finally,  $(\mathbb{N}, \emph{min})$ is a band (the extended
\emph{min} is still idempotent and associative) and the distributivity axioms also hold
for $\mathbb{N}_{\bot}$.

@@

As discussed in Chapter~\ref{sec:applications}, the lambda abstraction rule for coeffect systems
differs from the rule for effect systems in that it does not delay all context demands.
In case of implicit parameters (Section~\ref{sec:applications-flat-impl}), the demands
can be satisfied either by the call site or by the declaration-site. In case of dataflow and
liveness, the rule discussed in Section~\ref{sec:flat-calculus} reintroduces similar ambiguity
because it allows multiple valid typing derivations.

Furthermore, the semantics of context-aware languages in Chapter~\ref{sec:applications} and
also in Chapter~\ref{ch:semantics} is defined over \emph{typing derivation} and so the meaning
of a program depends on the typing derivation chosen. In this section, we specify how to choose
the desired \emph{unique} typing derivation in each of the coeffect systems we consider.

The most interesting case is that of implicit parameters. For example, consider the following
program written using the coeffect calculus with implicit parameter extensions:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~f~= (\kvd{let}~\ident{?x}=42~\kvd{in}~(\lambda y.~\ident{?x}))~\kvd{in} \\[-0.2em]
\kvd{let}~\ident{?x}=666~\kvd{in}~f~0 \\[-0.2em]
\end{array}
\end{equation*}

\noindent
There are two possible typings allowed by the typing rules discussed in Section~\ref{sec:flat-calculus-types}
that lead to two possible meanings of the program -- evaluating to $1$ and $2$, respectively:
%
\begin{itemize}
  \item $f : \ident{num} \xrightarrow{\cclrd{\emptyset}} \ident{num}$ -- in this case, the value
    of \ident{?x} is captured from the declaration-site and the program produces $1$.
  \item $f : \ident{num} \xrightarrow{\cclrd{\{ \ident{?x} \}}} \ident{num}$ -- in this case, the
    parameter \ident{?x} is required from the call site and the program produces $2$.
\end{itemize}
%
The coeffect calculus intentionally allows both of the options, acknowledging the fact that the
choice needs to be made for each individual concrete context-aware programming language. In the
above case, one typing derivation represents dynamic binding and the other static binding, but
more subtleties arise when the nested expression uses multiple implicit parameters.

In this section, we discuss the specific choices of typing derivation for implicit parameters,
dataflow and liveness. We use the fact that the coeffect calculus uses Church-style syntax for
lambda abstraction giving a type annotation for the bound variable. This does not affect the
handling of coeffects (those are not defined by the type annotation), but it lets us prove
\emph{uniqueness of typing}; a theorem showing that we define a \emph{unique} way of assigning
coeffects to otherwise well-typed programs.

@@

For implicit parameters we choose to follow the behaviour implemented by Haskell~\cite{app-implicit-parameters}
where function abstraction captures all parameters that are statically available at the declaration-site
and places all other demands on the call site. For the example above, this means
that the body of $f$ captures the value of \ident{?p} available from the declaration-site
and $f$ will be typed as a function requiring no parameters (coeffect $\cclrd{\emptyset}$). The
program thus evaluates to a numerical value $42$.

To express this behaviour formally, we extend the coeffect type system to additionally track
implicit parameters that are currently in static scope. The typing judgement becomes:
%
\begin{equation*}
\coctx{\Gamma;\cclrd{\Delta}}{\cclrd{r}} \vdash e : \tau
\end{equation*}
%
Here, $\cclrd{\Delta}$ is a set of implicit parameters that are in scope at the declaration-site.
The modified typing rules are shown in Figure~\ref{fig:flat-resolve-impl}. The rules (\emph{var}),
(\emph{const}), (\emph{app}) and (\emph{let}) are modified to use the new typing judgement, but they
simply propagate the information tracked by $\cclrd{\Delta}$ to all assumptions. The (\emph{param})
rule also remains unchanged -- the implicit parameter access is still tracked by the coeffect
$\cclrd{r}$ meaning that we still allow a form of dynamic binding (the parameter does not have
to be in static scope).

@@

\paragraph{Properties.}
If a program written in a coeffect language with implicit parameters is well-typed in a type
system presented in Figure~\ref{fig:flat-resolve-impl} then this identifies the unique preferred
derivation for the program. We use this unique typing derivation to give the semantics of coeffect
language with implicit parameters in Chapter~\ref{ch:semantics} and we also implement this algorithm
as discussed in Chapter~\ref{ch:impl}.

The type system is more restrictive than the fully general one and it reject certain programs that
could be typed using the more general system. This is expected -- we are restricting the fully
general coeffect calculus to match the typing and semantics of implicit parameters as known from
Haskell.

@@

\noindent
Finally, we note that unique typing derivations obtained using the type system given in
Figure~\ref{fig:flat-resolve-impl} are valid typing derivation under the original flat coeffect
type system in Figure~\ref{fig:flat-types}.

@@

When reducing the expression $(\lambda x.e_1)~e_2$ using the call-by-name strategy, the
sub-expression $e_2$ is substituted for all occurrences of the variable $v$ in an expression $e_1$.
As discussed in Section~\ref{sec:flat-syntax-props}, the call-by-name strategy does not \emph{in
general} preserve the type of a terms in coeffect calculi, but it does preserve the typing in two
interesting cases.

>>>>

In this chapter, we develop a \emph{flat coeffect calculus} that provides a type system for
tracking per-context properties of context-aware programming languages. The \emph{coeffect
calculus} captures the shared properties of such languages. It is parameterized by a \emph{flat
coeffect algebra} and can be instantiated to track implicit parameters, liveness and number of
required past values in dataflow languages.
To capture contextual properties in full generality, the flat coeffect calculus permits multiple
valid typing derivations for a given term. To resolve the ambiguity arising from such generality,
each concrete context-aware language is also equipped with an algorithm for choosing a unique
typing derivation. This allows us to explore the language design landscape, while still following
the usual scoping rules for languages with established approaches (e.g.~implicit parameters in
GHC).

@@

\item The coeffect calculus permits multiple typing derivations due to the ambiguity inherent
  in contextual lambda abstraction. Each concrete context-aware language based on the
  coeffect calculus must specify how such ambiguities are to be resolved. We give the procedures for
  choosing unique typing derivations for our three examples (Section~\ref{sec:flat-unique}).

@@

Despite the differences, the systems fit the same framework. This becomes apparent when
we consider the categorical structure (Chapter~\ref{ch:semantics}). Rather than starting
from the categorical semantics, we first explain how the systems can be unified syntactically
(Section~\ref{sec:flat-calculus-lambda}) and then provide the semantics as an additional
justification.

The development in this chapter can be seen as a counterpart to the well-known development of
\emph{effect systems} \cite{effects-gifford}. Chapter~\ref{ch:semantics} then links \emph{coeffects}
with \emph{comonads} in the same way in which effect systems have been linked with monads
\cite{monad-notions}. The syntax and type system of the flat coeffect calculus follows
a similar style to that of effect systems \cite{effects-polymorphic,effects-talpin-et-al}, but differs
in the structure of lambda abstraction as discussed briefly here and in Section~\ref{sec:applications-structure-lam}
(the relationship with monads is further discussed in Section~\ref{sec:semantics-related}).

@@

We enrich types and typing judgements with coeffect annotations $\cclrd{r}, \cclrd{s}, \cclrd{t}$;
typing judgements are written as $\coctx{\Gamma}{\cclrd{r}}\vdash e:\tau$. The expressions of
the calculus are those of the $\lambda$-calculus with \emph{let} binding. We also include a type
\ident{num} as an example of a concrete base type with numerical constants written as $n$:
%
\begin{equation*}
\begin{array}{rcl}
e &::=& x \sep n \sep \lambda x:\tau.e \sep e_1~e_2 \sep \kvd{let}~x = e_1~\kvd{in}~e_2\\
\tau &::=& \ident{num} \sep \tau_1 \xrightarrow{\cclrd{r}} \tau_2
\end{array}
\end{equation*}
%
Note that the lambda abstraction in the syntax is written in the Church style and requires a type
annotation. This will be used in Section~\ref{sec:flat-unique} where we discuss how to find a
unique typing derivation for context-aware computations. Using Church style lambda abstraction,
we can directly focus on the more interesting problem of finding unique \emph{coeffect annotations}
rather than solving the problem of type reconstruction.

@@

\begin{definition}
A \emph{\cclrd{flat coeffect algebra}} $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$ is a set
$\C$ together with elements $\cunit, \czero \in \C$, binary relation $\cleq$ and binary operations
$\cseq, \cpar, \czip$ such that $(\C, \cseq, \cunit)$ is a monoid, $(\C, \cpar, \czero)$ is an
idempotent monoid, $(\C, \czip)$ is a band (idempotent semigroup) and $(\C, \cleq)$ is a pre-order.
That is, for all $r,s,t\in \C$:
%
\begin{equation*}
\begin{array}{ccc}
r \;\cseq\; (s \;\cseq\; t) = (r \;\cseq\; s) \;\cseq\; t &&
\cunit \;\cseq\; r = r = r \;\cseq\; \cunit
\\
r \;\cpar\; (s \;\cpar\; t) = (r \;\cpar\; s) \;\cpar\; t &r\; \cpar\; r = r~~~~&
\czero \;\cpar\; r = r = r \;\cpar\; \czero
\\
r \;\czip\; (s \;\czip\; t) = (r \;\czip\; s) \;\czip\; t &r\; \czip\; r = r~~~~&~
\\
\textnormal{if}~~r\; \cleq\; s ~~\textnormal{and}~~s\; \cleq\; t~~\textnormal{then}~~r\; \cleq\; t &
t\; \cleq\; t \\[0.5em]
\end{array}
\end{equation*}
%
In addition, the following distributivity axioms hold:
\begin{align*}
\quad (\cclrd{r}\, \cpar\, \cclrd{s}) \;\cseq\; \cclrd{t} & = (\cclrd{r} \,\cseq\, \cclrd{t}) \;\cpar\; (\cclrd{s}\, \cseq\, \cclrd{t}) \\
\quad \cclrd{t} \;\cseq\; (\cclrd{r}\, \cpar\, \cclrd{s}) & = (\cclrd{t} \,\cseq\, \cclrd{r}) \;\cpar\; (\cclrd{t}\, \cseq\, \cclrd{s})
\end{align*}
\end{definition}

@@

\begin{itemize}
\item The monoid $(\C, \cseq, \cunit)$ represents \emph{sequential} composition of (semantic)
functions. The monoid axioms are required in order to form a category structure in the
semantics (Section~\ref{sec:semantics-theory}).

@@

\noindent
It is worth noting that, in some of the systems, the operators $\cpar$ and $\czip$ are the least
upper bound and the greatest lower bounds of a lattice. For example, in dataflow computations, they
are \emph{max} and \emph{min} respectively. However, this duality does not hold for implicit parameters.

\paragraph{Semiring.}
The $\cseq$ and $\cpar$ operations of the flat coeffect algebra together with the $\cunit$ and
$\czero$ values resemble semiring. As required by semiring laws, both of the operations are 
associative; $\cunit$ and $\czero$ are the units of $\cseq$ and $\cpar$, respectively, and the
distributivity laws also hold. The two additional laws of semiring that are not required by a 
flat coeffect algebra are symmetry of $\cseq$ and annihilation:
%
\begin{align*}
\czero \, \cseq \, \cclrd{r} = \czero = \cclrd{r} \, \cseq \, \czero \qquad \cclrd{r}\,\cpar\,\cclrd{s} = \cclrd{s}\,\cpar\,\cclrd{r}
\end{align*}
%
The symmetry of $\cseq$ holds for all our examples, but we do not include it so that we do not
unnecessarily restrict the system. The annihilation holds for some of the examples (liveness
and optimized dataflow in Section~\ref{sec:flat-calculus-examples}), but not for all of them.
In particular, it does not hold for implicit parameters $\cclrd{\emptyset}\,\cclrd{\cup}\,\cclrd{r}\neq\cclrd{\emptyset}$
and so requiring it would lead to a more restrictive structure than is desirable.

\paragraph{Ordering.}
The flat coeffect algebra includes a pre-order relation $\cleq$. This will be used to introduce
subcoeffecting and subtyping in Section~\ref{sec:flat-exts-sub}, but we make it a part of the flat
coeffect algebra, as it will be useful for characterization of different kinds of coeffect calculi.
When the idempotent monoid $(\C, \cpar, \czero)$ is also commutative (\ie~forms a
semi-lattice), the $\cleq$ relation can be defined as the ordering of the semi-lattice:
%
\begin{equation*}
\cclrd{r} \;\cleq\; \cclrd{s} \;\Longleftrightarrow\; \cclrd{r} \;\cpar\; \cclrd{s} \;=\; \cclrd{s}
\end{equation*}
%
This definition is consistent with all three examples that motivate flat coeffect calculus, but
it cannot be used with the structural coeffects (Chapter~\ref{ch:structural}), where it fails for
the bounded reuse calculus. For this reason, we choose not to use it for flat coeffect calculus
either.

@@

To guide understanding of the system, we also show a non-syntax-directed (\emph{sub}) rule for
subcoeffecting. The rule states that an expression with context demands $\cclrd{r'}$ can be
treated as an expression with greater context demands $\cclrd{r}$. We return to subcoeffecting,
subtyping and additional constructs such as pairs in Section~\ref{sec:flat-exts}. When
discussing the procedure for choosing unique typing in Section~\ref{sec:flat-unique}, we consider
only the syntax-directed part of the system.

\subsection{Understanding flat coeffects}
\label{sec:flat-calculus-understanding}

Before proceeding, let us clarify how the typing judgements should be understood. The coeffect
calculus can be understood in two ways discussed in this and the next chapter. As a type system
(Chapter~\ref{ch:flat}), it provides analysis of context dependence. As a semantics
(Chapter~\ref{ch:semantics}), it specifies how context is propagated. These two readings provide
different ways of interpreting the judgement $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ and the
typing rules used to define it.

@@

\noindent
For example, using the syntactic reading, the operators $\czip$ and $\cpar$ represent \emph{merging}
and \emph{splitting} of context demands -- in the (\emph{abs}) rule, $\czip$ appears in the assumption
and the combined context demands of the body are split between two positions in the conclusions;
in the (\emph{app}) rule, $\cpar$ appears in the conclusion and merges two context demands
from the assumptions.

@@

\noindent
For simplicity, assume that all parameters have the same type $\ident{num}$ and so the annotations only
track sets of names. The definition uses a set union for all three operations. Both variables and
constants are annotated with $\emptyset$ and the ordering is defined by $\subseteq$. The
definition satisfies the flat coeffect algebra axioms because $(S, \cup, \emptyset)$ is an idempotent,
commutative monoid. The language has additional syntax for defining an implicit parameter and for
accessing it, together with associated typing rules:
%
\begin{equation*}
\quad~ e ~::=~ \ldots \sep \ident{?p} \sep \kvd{let}~\ident{?p}~=~e_1~\kvd{in}~e_2
\end{equation*}
\begin{equation*}
\tyrule{param}
  { ~ }
  { \coctx{\Gamma}{\cclrd{ \{ \ident{?p} \} }} \vdash \ident{?p} : \ident{num} }
\end{equation*}
\begin{equation*}
\tyrule{letpar}
  { \coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 &
    \coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_2}
  {\coctx{\Gamma}{\cclrd{r} \;\cclrd{\cup\; (\cclrd{s} \setminus \{ \ident{?p}) \}} } \vdash \kvd{let}~\ident{?p}=e_1~\kvd{in}~e_2 : \tau_2 }
\end{equation*}
%
The (\emph{param}) rule specifies that the accessed parameter $\ident{?p}$ needs to be in the set of
required parameters $\cclrd{r}$. As discussed earlier, we use the same type $\ident{num}$ for all
parameters, but it is also possible to define a coeffect calculus that uses mappings from names to
types.

@@

As in Section~\ref{sec:applications-flat-live}, sequential composition $\cseq$ is modelled by
the meet operation $\sqcap$ and pointwise composition $\cpar$ is modelled by join $\sqcup$.
The two-point lattice induces a commutative, idempotent monoid. Distributivity
$(r \sqcup s) \sqcap t = (r \sqcap t) \sqcup (s \sqcap t)$ does not hold for \emph{every}
lattice, but it trivially holds for the two-point lattice used here.

@@

As required by the axioms, $(\mathbb{N}, +, 0)$ and $(\mathbb{N}, \mathit{max}, 0)$ form monoids
and $(\mathbb{N}, \mathit{min})$ forms a band. Note that dataflow is our first example where
$\cseq$ is not idempotent. The distributivity axioms require addition to distribute over maximum,
which is easy to see.

A simple dataflow language includes an additional construct $\kvd{prev}$ for accessing the
previous value in a stream with an additional typing rule that looks as follows:
%
\begin{equation*}
\quad~ e ~::=~ \ldots \sep \kvd{next}~e
\end{equation*}
\begin{equation*}
\tyrule{prev}
  { \coctx{\Gamma}{\cclrd{n}} \vdash e : \tau }
  { \coctx{\Gamma}{\cclrd{n+1}} \vdash \kvd{prev}~e : \tau }
\end{equation*}
%
As a further example that was not covered earlier, it is also possible to combine liveness analysis
and dataflow. In the above dataflow calculus, $0$ denotes that we require the current value of some
variable, but no previous values. However, for constants, we do not even need the current value.

\begin{example}[Optimized dataflow]
In optimized dataflow, context is annotated with natural numbers extended with the $-\infty$ element.
The flat coeffect algebra is $(\mathbb{N} \cup \{-\infty\}, +, \mathit{max}, \mathit{min}, 0, \bot, \leq)$
where $m + n$ is $-\infty$ whenever $m=-\infty$ or $n=-\infty$ and \emph{min}, \emph{max} treat $-\infty$ as the
least element.
\end{example}

\noindent
Note that $(\mathbb{N}\cup\{-\infty\}, +, 0)$ is a monoid for the extended definition of $+$; for the
bottom element $0 + \bot = \bot$ and for natural numbers $0 + n = n$. The structure
$(\mathbb{N}, \emph{max}, \bot)$ is also a monoid, because $\bot$ is the least element and so
$\mathit{max}(n, \bot) = n$. Finally,  $(\mathbb{N}, \emph{min})$ is a band and the distributivity 
axioms also hold for $\mathbb{N}\cup\{-\infty\}$.

@@

As discussed in Chapter~\ref{ch:applications}, the lambda abstraction rule for coeffect systems
differs from the rule for effect systems in that it does not delay all context demands.
In case of implicit parameters (Section~\ref{sec:applications-flat-impl}), the demands
can be satisfied either by the call site or by the declaration site. In case of dataflow and
liveness, the rule discussed in Section~\ref{sec:flat-calculus} reintroduces similar ambiguity
because it allows multiple valid typing derivations.

Furthermore, the semantics of context-aware languages in Chapter~\ref{ch:applications} and
also in Chapter~\ref{ch:semantics} is defined over \emph{typing derivations} and so the meaning
of a program depends on the typing derivation chosen. In this section, we specify how to choose
the desired \emph{unique} typing derivation in each of the coeffect systems we consider.

The most interesting case is that of implicit parameters. Consider the following
program written using the extended coeffect calculus:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{?x}=\num{1}\\[-0.2em]
\kvd{let}~f~= \lambda y.~\ident{?x}~\kvd{in} \\[-0.2em]
\kvd{let}~\ident{?x}=2~\kvd{in}~f~0 \\[-0.2em]
\end{array}
\end{equation*}

\noindent
There are two possible typings allowed by the typing rules discussed in Section~\ref{sec:flat-calculus-types}
that lead to two possible meanings of the program -- evaluating to $1$ and $2$, respectively:
%
\begin{itemize}
  \item $f : \ident{num} \xrightarrow{\cclrd{\emptyset}} \ident{num}$ -- in this case, the value
    of \ident{?x} is captured from the declaration site and the program produces $1$.
  \item $f : \ident{num} \xrightarrow{\cclrd{\{ \ident{?x} \}}} \ident{num}$ -- in this case, the
    parameter \ident{?x} is required from the call site and the program produces $2$.
\end{itemize}
%
The coeffect calculus intentionally allows both options, acknowledging the fact that the
choice needs to be made for each individual concrete context-aware programming language. In the
above case, one typing derivation represents dynamic binding and the other static binding, but
more subtleties arise when the nested expression uses multiple implicit parameters.

In this section, we discuss the specific choices of typing derivation for implicit parameters,
dataflow and liveness. We use the fact that the coeffect calculus uses Church style syntax for
lambda abstraction giving a type annotation for the bound variable. This does not affect the
handling of coeffects (those are not defined by the type annotation), but it lets us prove
\emph{uniqueness of typing}; a theorem showing that we define a \emph{unique} way of assigning
coeffects to otherwise well-typed programs.

@@

For implicit parameters we choose to follow the behaviour implemented by GHC~\cite{app-implicit-parameters}
where function abstraction captures all parameters that are statically available at the declaration site
and places all other demands on the call site. For the example above, this means
that the body of $f$ captures the value of \ident{?p} available from the declaration site
and $f$ will be typed as a function requiring no parameters (coeffect $\cclrd{\emptyset}$). The
program thus evaluates to a numerical value $1$.

To express this behaviour formally, we extend the coeffect type system to additionally track
implicit parameters that are currently in static scope. The typing judgement becomes:
%
\begin{equation*}
\coctx{\Gamma;\cclrd{\Delta}}{\cclrd{r}} \vdash e : \tau
\end{equation*}
%
Here, $\cclrd{\Delta}$ is a set of implicit parameters that are in scope at the declaration site.
The modified typing rules are shown in Figure~\ref{fig:flat-resolve-impl}. The rules (\emph{var}),
(\emph{const}), (\emph{app}) and (\emph{let}) are modified to use the new typing judgement, but they
simply propagate the information tracked by $\cclrd{\Delta}$ to all assumptions. The (\emph{param})
rule also remains unchanged -- the implicit parameter access is still tracked by the coeffect
$\cclrd{r}$ meaning that we still allow a form of dynamic binding (the parameter does not have
to be in static scope).

@@

\paragraph{Properties.}
If a program written in a coeffect language with implicit parameters is well-typed in the type
system presented in Figure~\ref{fig:flat-resolve-impl} then this identifies the unique preferred
derivation for the program. We use this unique typing derivation to give the semantics of a coeffect
language with implicit parameters in Chapter~\ref{ch:semantics} and we also implement this algorithm
as discussed in Chapter~\ref{ch:impl}.

The type system is more restrictive than the fully general one and it rejects certain programs that
could be typed using the more general system. This is expected -- we are restricting the fully
general coeffect calculus to match the typing and semantics of implicit parameters as known from
Haskell.

@@

\noindent
Finally, we note that unique typing derivations obtained using the type system given in
Figure~\ref{fig:flat-resolve-impl} are valid typing derivations under the original flat coeffect
type system in Figure~\ref{fig:flat-types}.

@@

When reducing the expression $(\lambda x.e_1)~e_2$ using the call-by-name strategy, the
sub-expression $e_2$ is substituted for all occurrences of the variable $v$ in an expression $e_1$.
As discussed in Section~\ref{sec:flat-syntax-props}, the call-by-name strategy does not \emph{in
general} preserve the type of a term in coeffect calculi, but it does preserve the typing in two
interesting cases.
